### 2016-01-09 papers/CAFe/HIPS-paper/coarray-extensions.tex
Finished first cut by adding to subsection on integration with standard CAF.

In addition, CAFe subimage tasks may also be defined by the body of a Fortran do
concurrent construct and executed on a subimage.  For example, the code segment,
\begin{verbatim}
do concurrent(i=1:N-1)  [[device]]
 begin block
   real :: w = 0.67
   S(i) = (1-w)*T(i) + w*(T(i-1)+T(i+1)/2)
 end block
end do
\end{verbatim}
will execute all of the code within the \texttt{do concurrent} construct on \texttt{device}.

In Fortran, the execution semantics of a \texttt{do concurrent} loop is that the loop
iterates can be executed concurrently in any order.  Note the similarities between this
example and the OpenMP pragma \texttt{!\$omp parallel do, private(w)}.  The use of the
Fortran block construct to declare private variables allows a natural and clear way to
distinguish between variables that are private to a ``thread'' (although in the example
above, the ``private'' distinction for the variable \texttt{w} is unnecessary because it
is a constant for all loop iterations).

(NOTE FOR CRAIG: need to read the standard to see if a block construct can be used in this
way. ALSO, there is a lot more work here to think about how do concurrent merges with
OpenMP and OpenACC with regards to synchronization, barriers, teams, atomics, , critical
sections, ...)

\subsection{Integration with Coarray Standard Features}

Want to address: Barriers and teams primarily

%\input{coarray-comparison}

%%A team of images is a set of images that can readily execute independently of other images (from N2007)

%%An allocated allocatable coarray is established in the team in which it was allocated

%%\subsection{Execution Semantics and Memory Management}

%%This section describes the hierarchical memory layout and how memory consistency between the 3-levels of memory is maintained.  It describes what is the compiler's responsibility and what is the programmer's responsibility.

%% Completion of a do concurrent construct indicates that all executing threads (if a threading model is used by the compiler) have completed and that all memory on the executing subimage is in a consistent state.

%%The {\tt HALO} function returns a copy of the 3x3 region of {\tt a} and its surrounding neighbors.  The {\tt convolve} function is free of race conditions because of the copy semantics of {\tt HALO} and the implied synchronization of the {\tt CONCURRENT} attribute, whereby no output variables can be updated before all threads have completed execution.  In addition, while any thread may load from an extended region about \emph{its} element with the {\tt HALO} function, it may only store into its own elemental location.

%% A note from a conversation with Matt regarding memory management
%%

%The Fortran language has much tighter restrictions on aliasing than does C.
%So unless a variable has the pointer or target attribute, it cannot be aliased.
%Thus the compiler is able to aggressively optimized for memory movement between
%the CPU and accelerator.  However, because the design philosophy of coarrays
%is that memory transfer between images can be expensive, the programmer must
%explicitly transfer memory between images with explicit syntax with square
%bracket notation, i.e., $a[1] = a[2]$.  So we allow the compiler to manage
%memory within an image but require the use of {\tt halo\_exchange} for the
%transfer of halo memory between images.

%Fortran currently supports:

%1. Array syntax: e.g., C = A + B, where A, B, C are arrays.  Note that this is implicitly a loop structure, but that no loop indices need be provided.  Also the programmer need not specify where in memory these arrays reside.  Thus this high level syntax allows the compiler more freedom in both memory placement (even across distributed memory nodes) and in runtime code execution (individual array element may be computed by different hardware threads).  This is a simple example and it is a research question as to what compiler directives would be useful for memory placement and other directions to the compiler for efficient code generation.

%2. Pure procedures:  Fortran has syntax for specifying procedures that have no side effects during execution.  Specifying code that is side-effect free code is important information to provide to the compiler so that it can generate efficient multi-threaded code.

%3. Pure elemental procedures: Fortran has syntax for specifying procedures that take only scalar arguments, but may be applied across array elements.  Elemental procedures are ideal for writing code to be executed within a hardware thread.  They resemble OpenCL kernels, but are simpler because they leave all indexing up to the compiler.

%We have determined that additional syntax is needed, in addition to the three language features described above, to allow programmers the ability to express code in Fortran to be targeted for multi-threaded hardware architectures like GPUs.  This additional syntax is provided by functions that return a copy of a small region of memory surrounding an array element (as seen within an elemental procedure) and with functions for thread synchronization.  This additional syntax will allow pure procedures to perform stencil and other convolution-like operations on a copy of memory, synchronize, then store the computed results back to the array element associated with the given thread.

%In addition Fortran has syntax like the target attribute the specifies when variables can be aliased.  This allows for much easier program analysis as the compiler knows that ordinary variables cannot be aliased.  Fortran also has excellent facilities for interoperability with C so that programming in a mixed language environment is easily accomplished, including interoperability with native Fortran arrays.

%
% This could be part of conclusions
%

%%Note that the use of halo cells is the normal way that large and complex MPI and CAF programs are implemented.  LOPe proposes to formalize this common pattern into the Fortran language allowing the compiler access to this information in order to spread computation over more hardware resources, improve performance, and to reduce complexity for the programmer.

>>>>

In addition, CAFe subimage tasks may also be defined by the body of a Fortran do
concurrent construct and executed on a subimage.  For example, the code segment,
\begin{verbatim}
do concurrent(i=1:N-1)  [[device]]
 begin block
   real :: w = 0.67
   S(i) = (1-w)*T(i) + w*(T(i-1)+T(i+1)/2)
 end block
end do
\end{verbatim}
will execute all of the code within the \texttt{do concurrent} construct on \texttt{device}.
Execution of a \texttt{do concurrent} loop on a subimage acts \emph{as if} the loop were
extracted as a outlined function and the subsequent task executed on a subimage.  Thus
there is an implicit barrier at the end of the loop that blocks the hosting image
from continuing execution until \emph{all} of the iterates of the loop have finished.

In Fortran, the execution semantics of a \texttt{do concurrent} loop is that loop
iterates can be executed concurrently in any order.  Note the similarities between this
example and the OpenMP pragma \texttt{!\$omp parallel do, private(w)}.  The use of the
Fortran block construct to declare private variables allows a natural and clear way to
distinguish between variables that are private to a ``thread'' (although in the example
above, the ``private'' distinction for the variable \texttt{w} is unnecessary because it
is a constant over all loop iterations).

The comparison with OpenMP also suggests that subimage execution may not be restricted to
a single hardware execution unit, but may indeed be executed by a team of threads in a
single (shared) memory address space.  This association with a team of threads
will be exploited below when considering a team of images.

\begin{comment}
  (NOTE FOR CRAIG: need to read the standard to see if a block construct can be used in this
way. ALSO, there is a lot more work here to think about how do concurrent merges with
OpenMP and OpenACC with regards to synchronization, barriers, teams, atomics, , critical
sections, ...)
\end{comment}

\subsection{Integration with Standard Coarray Features}

The new CAFe features introduced above work in concert with the existing 2008 Fortran
coarray constructs and with the new parallel features to appear in the next Fortran
standard\cite{TS:18508} (referred to here as Fortran 2015).  This section describes
in more detail how the proposed extensions of CAFe interact with the parallel
language features of Fortran 2015.

\subsubsection{Fortran 2008 Standard}

Perhaps the most important of the 2008 parallel constructs are those involving the
control of image execution, including \texttt{sync all} and \texttt{sync images}
statements.  A \texttt{sync all} statement implements a barrier synchronizing all CAF
images while a \texttt{sync images} statement synchronizes a specified set of images.
Neither of these synchronization primitives affect tasks potentially executing concurrently
on a subimage.  Once subimage tasks begin execution they run to completion from the perspective
of the hosting image.

Other features of 2008 include the \texttt{critical} construct and the \texttt{lock} and
\texttt{unlock} statements.  The \texttt{critical} construct limits the execution of a
block of code to one image at a time, while lock variables provide a facility for atomic
operations.  Both of these features are defined in terms of normal CAF images so subimages
do not directly participate unless the parent (or host) of a subimage does.

\subsubsection{Fortran 2015 Standard}

The most important concept introduced by the 2015 standard is that of a team of images.
However, subimages may not be included within a team of standard CAF images.  However, if one
considers the comparison of the execution of a \texttt{do concurrent} block on a subimage
with a team of OpenMP threads (as described above), the subimage could be considered as
a team of one member that \emph{may} be executing its tasks on a team of shared memory threads.
This allows allows the use of a \texttt{sync team} statement to synchronize with
possibly executing threads.

Thus we introduce another CAFe function, \texttt{GET\_SUBIMAGE\_TEAM()}, that returns a
variable representing a fully-formed (established) team consisting of only one team
member, the current subimage.  If this function is called outside the context of a task
executing on a subimage it will return only a team consisting of the current image,
\texttt{this\_image()}.  Synchronization across possible subimage threads is then possible
using the statement
\begin{verbatim}
  sync team (GET_SUBIMAGE_TEAM())
\end{verbatim}
This allows implementation of the technique of double buffering within a subimage task
so that all of the \emph{potential} subimage threads will have completed operating on the
temporary buffer before it is copied back to the primary array variable.

Fortran 2015 also introduces the concept of events.  An event is a Fortran type that
can be posted to by one image and waited for by another image.  CAFe relaxes the
constraint that events must be coarrays so that events can be used locally within
a single image to control concurrent execution of multiple subimage tasks.  For example,
the following code segment shows how two tasks can be run concurrently on two separate
subimages:
\begin{verbatim}
 type(event_type) :: evt
 call update_ocn [[dev1, WITH_EVENT=evt]]
 call update_atm [[dev2, WITH_EVENT=evt]]
 event wait (evt, until_count=2)
\end{verbatim}

In this example, the syntax \texttt{[[dev1, WITH\_EVENT=evt]]} indicates that the called
subroutine is to be run on \texttt{dev1} and that the event \texttt{evt} will receive
notification (an implicit posting) once the task has completed.  The \texttt{event wait}
statement will cause execution on the hosting image to wait until \emph{both} tasks have
completed.  One may even run multiple tasks on the \emph{same} device using events,
however unless the subimage hardware allows concurrent execution of multiple tasks, only
one task will be executed at a time.

Fortran 2015 also provides new intrinsic procedures for atomic and collective operations.
At this time we do not consider atomic operations for usage with the CAFe extensions.
Atomic variables and associated synchronization constructs place too great of requirements
on the executing software and hardware environment.  For example, the Open Community
Runtime environment does not provide for atomic variables as the restriction
helps OCR to better scale on a wider range of parallel computers.

Collective operations, however, could potentially be very useful and fit nicely within
CAFe concepts.  For example, a team consisting of hosted subimages could be formed and
then data broadcasted to team members and reduction operations such as \texttt{co\_add}
performed.  However since all team members of the current team must participate in
collective operations \emph{and} since CAFe restricts subimages from being part of the
same team with regular images, it makes it impossible for an image to participate with a
team of its hosted subimages in collective operations.  So for now, collectives are
excluded from CAFe.
